Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-09-11 pot/LaTeX/skewnormal.tex
results compilation

In this paper, we present a model that has marginal distributions with flexible tails, demonstrates asymptotic dependence for small $\bh$, and has computation on the order of Gaussian models for large space-time datasets.
Specifically, our contribution is to incorporate thresholding and random spatial partitions using a multivariate skew-$t$ distribution.
The advantage of using a thresholded model as opposed to a non-thresholded model is that is allows for the tails of the distribution to inform the predictions in the tails \citep{DuMouchel1983}.
The random spatial partition alleviates the long-range spatial dependence seen by the skew-$t$.

\section{Statistical model}\label{s:model}
Let $Y_t(\bs)\in \calR$ be the observed value at location $\bs$ and timepoint $t$.
To avoid bias in estimating tail parameters, we model censored data
\beq\label{Yt}
  \Yt_t(\bs) = \left\{
          \begin{array}{ll}
            Y_t(\bs) & Y_t(\bs)>T \\
            T & Y_t(\bs)\le T
          \end{array}
        \right.
\eeq
where $T$ is a pre-specified threshold.
Then, assuming the full data follow a skew-$t$ distribution, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Skew-$t$ process}\label{s:data}
We assume the data can be modeled as skew-$t$.
\citet{Zhang2010} show that the skew-$t$ can be written as the hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N_{(0, \infty)}(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.3), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, and \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail, controlled by both the skewness parameter $\alpha$ and the degrees of freedom $2a$, and its joint distribution is a multivariate skew-$t$ distribution.

@@

Another popular measure of extremal dependence is the $\chi$ statistic.
The $\chi$ coefficient for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c)
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(\bh) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(\bh) > c)
\end{align*}
where $\bh = ||\bs_1 - \bs_2||$.
If $\chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
For Gaussian processes, $\chi(\bh) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the process described in Section \ref{s:data}, $\chi(\bh) > 0$ \citep{Padoan2011}.

\subsection{Random daily partition}\label{s:part}
One problem with the skew-$t$ distribution is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this is given in Appendix A.2.

\subsection{Hierarchical model}\label{s:hier}
Conditioned on $z_{tk}(\bs) \iid $ $N(0, \sigma^2_{tk})$, $\sigma^2_{tk}(\bs) \sim IG(a, b)$, and $P_tk$, the marginal distributions are skew-$t$ and the joint distribution within a partition is multivariate skew-$t$. However, we do not fix the partitions, they are treated as unknown and updated in the MCMC.
Thus, standard geostatistical methods can be used to fit the model, and predictions can be made by Kriging at unobserved locations.
We model this with a Bayesian hierarchical model as follows.
Let $w_{t1}, \ldots, w_{tK}$ be a set of daily spatial knots in a spatial domain of interest, $\calD$, so that
\begin{align*}
 P_{tk} = \{\bs : k = \argmin_\ell || \bs_t - w_{t\ell} ||\}.
\end{align*}
Then
\begin{align}
   Y_t(\bs) \mid \Theta, z_{t}(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma
   _t(\bs) v_t(\bs) \label{eq:hier}\\
   z_t(\bs) = z_{tk} \text{ if } \bs \in P_{tk}\\
   \sigma^2_{t}(\bs) = \sigma^2_{tk} \text{ if } \bs \in P_{tk}
   z_{tk} \mid \sigma^2_{tk} &\sim N_{(0, \infty)}(0, \sigma^2_{tk})\\
   \sigma^2_{tk} &\iid IG(\alpha, \beta)\\
   v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma)\\
   \alpha &\sim N(0, 10)\\
   w_{tk} &\sim Unif(\calD)
\end{align}
where $\Theta = \{w_{t1}, \ldots, w_{tK}, \beta, \sigma_t, \alpha, \lambda, \rho, \nu \}$; \mbox{$k = \argmin_\ell ||\bs - \bw_\ell ||$}; and $\Sigma$ is a \Matern covariance matrix with variance one, spatial range $\rho$, smoothness $\nu$.

@@

\subsection{Imputation}\label{s:impute}
We can use Gibbs sampling to update $Y_t(\bs)$ for observations that are below $T$, the thresholded value. Given $\Theta$, $Y_t(\bs)$ has truncated normal full conditional with these parameter values.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}(\mu(\bs), \Sigma)$

\subsection{Parameter updates}\label{s:params}
To update $\Theta$ given the current value of the complete data $\bY_1, \ldots, \bY_T$, we use a standard Gibbs updates for all parameters except for the knot locations which are done using a Metropolis update.
See Appendix A.1 for details regarding Gibbs sampling.

\subsection{Spatial prediction}\label{s:pred}
Given $\bY_t$ the usual Kriging equations give the predictive distribution for $Y_t(\bs^*)$ at prediction location $(\bs^*)$

@@

$M = 50$ data sets are generated for each design.
For each data set we fit the data using
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=1$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=1$ knots, $T=q(0.90)$
  \item skew-$t$ marginal, $K=5$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=5$ knots, $T=q(0.9)$
\end{enumerate}
where $q(0.9)$ is the 90th sample quantile of the data.
The design matrix $\bX$ includes on the intercept with a prior of $\beta \sim \text{N}(0, 10)$.
The spatial covariance parameters have priors $\log(\nu) \sim \text{N}(-1.2, 1)$, $\gamma \sim \text{U}(0, 1)$, $\log(\rho) \sim \text{N}(-2, 1)$.
The skewness parameter has prior $\alpha \sim \text{N}(0, 2)$.
The residual variance terms have priors $\sigma^2_t(\bs) \sim \text{IG}(0.1, 0.1)$.

@@

\section{Data analysis}\label{s:analysis}
To illustrate this method, we consider the daily maximum 8-hour ozone measurements for July 2005 at 735 Air Quality System (AQS) monitoring sites in the eastern United States as the response.
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
We fit the model using Gaussian and skew-$t$ marginal distributions, $K=1, 5, 10, 15$ partitions, with $Y(\bs)$ censored at $T = 0, 50, 75, 90$ ppb as described in Section \ref{s:data}.
We also include a max-stable analysis using the method by ????
All methods assume the location can be expressed as
\begin{align}
  \mu_t(\bs) = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs).
\end{align}
To explore the extremal dependence both over space and time, we plot $\chi(\bs)$ and $\chi(t)$.
For each model, Brier scores and quantile scores were were averaged over all sites and days to obtain a single quantile score for each dataset.
At a particular threshold or quantile level, the model that fits the best is the one with the lowest score.

\subsection{Results}\label{s:results}

>>>>

In this paper, we present a model that has marginal distributions with flexible tails, demonstrates asymptotic dependence for observations at sites that are near to one another, and has computation on the order of Gaussian models for large space-time datasets.
Specifically, our contribution is to incorporate thresholding and random spatial partitions using a multivariate skew-$t$ distribution.
The advantage of using a thresholded model as opposed to a non-thresholded model is that is allows for the tails of the distribution to inform the predictions in the tails \citep{DuMouchel1983}.
The random spatial partition alleviates the long-range spatial dependence seen by the skew-$t$.

The paper is organized as follows.
In section \ref{s:model}, we describe the skew-$t$ process with partitioning and thresholding.
The computing is described in section \ref{s:comp}.
In section \ref{s:simstudy}, we present a simulation study that examines the predictive capabilities of this model compared with a na\:{i}ve Gaussian method.
We then compare our method to Gaussian and max-stable methods with a data analysis of ozone measurements from the eastern US in section \ref{s:analysis}.
The final section provides brief discussion and direction for future research.

\section{Statistical model}\label{s:model}
Let $Y_t(\bs)\in \calR$ be the observed value at location $\bs$ and timepoint $t$.
To avoid bias in estimating tail parameters, we model censored data
\beq\label{Yt}
  \Yt_t(\bs) = \left\{
          \begin{array}{ll}
            Y_t(\bs) & Y_t(\bs)>T \\
            T & Y_t(\bs)\le T
          \end{array}
        \right.
\eeq
where $T$ is a pre-specified threshold.
Then, assuming the full data are observations from a skew-$t$ process, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
We assume the data can be modeled using a skew-$t$ distribution.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N_{(0, \infty)}(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.3), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

@@

Another measure of extremal dependence is the $\chi$ statistic.
The $\chi$ statistic for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c),
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(\bh) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(\bh) > c)
\end{align*}
where $\bh = ||\bs_1 - \bs_2||$.
If $\chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
For Gaussian processes, $\chi(\bh) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the process described in Section \ref{s:data}, $\chi(\bh) > 0$ \citep{Padoan2011}.

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Huser2014} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.2.

\subsection{Hierarchical model}\label{s:hier}
Conditioned on $z_{tk}(\bs) \iid $ $N(0, \sigma^2_{tk})$, $\sigma^2_{tk}(\bs) \sim IG(a, b)$, and $P_tk$, the marginal distributions are skew-$t$ and the joint distribution within a partition is multivariate skew-$t$.
However, we do not fix the partitions, they are treated as unknown and updated in the MCMC.
We model this with a Bayesian hierarchical model as follows.
Let $w_{t1}, \ldots, w_{tK}$ be a set of daily spatial knots in a spatial domain of interest, $\calD$, so that
\begin{align*}
 P_{tk} = \{\bs : k = \argmin_\ell || \bs_t - w_{t\ell} ||\}.
\end{align*}
Then
\begin{align}
   Y_t(\bs) \mid \Theta, z_{t}(\bs) &= X_t(\bs) \beta + \alpha z_t(\bs) + \sigma
   _t(\bs) v_t(\bs) \label{eq:hier}\\
   z_t(\bs) &= z_{tk} \text{ if } \bs \in P_{tk}\\
   \sigma^2_{t}(\bs) &= \sigma^2_{tk} \text{ if } \bs \in P_{tk}\\
   z_{tk} \mid \sigma^2_{tk} &\sim N_{(0, \infty)}(0, \sigma^2_{tk})\\
   \sigma^2_{tk} &\iid IG(\alpha, \beta)\\
   v_t(\bs) \mid \Theta &\sim \Matern(0, \Sigma)\\
   \alpha &\sim N(0, 10)\\
   w_{tk} &\sim Unif(\calD)
\end{align}
where $\Theta = \{w_{t1}, \ldots, w_{tK}, \beta, \sigma_t, \alpha, \lambda, \rho, \nu \}$; \mbox{$k = \argmin_\ell ||\bs - \bw_\ell ||$}; and $\Sigma$ is a \Matern covariance matrix with variance one, spatial range $\rho$, smoothness $\nu$.

@@

We can use Gibbs sampling to update $Y_t(\bs)$ for censored observations that are below the threshold $T$.
After conditioning on $\alpha$, $z_t(\bs)$ and non-censored observations, $Y_t(\bs)$ has truncated normal full conditionals.
So we sample $Y_t(\bs) \sim N_{(-\infty, T)}(\mu(\bs), \Sigma)$.
After imputing the censored observations, we update the model parameters.
To update the model parameters, we use standard Gibbs updates for parameters when possible.
In the case Gibbs sampling is not possible, parameters are updated using a random-walk Metropolis Hastings algorithm.
See Appendix A.1 for details regarding Gibbs sampling and Metropolis update rules.
The final step of the computation is to use Bayesian Kriging to generate a predictive distribution for $Y_t(\bs^*)$ at prediction location $(\bs^*)$.
This step is similar to the imputation for censored observations except that the full conditionals are no longer truncated at $T$.

@@

$M = 50$ data sets are generated for each design.
For each data set we fit the data using
\begin{enumerate} \setlength{\itemsep}{-0.5em}
  \item Gaussian marginal, $K=1$ knots
  \item skew-$t$ marginal, $K=1$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=1$ knots, $T=q(0.80)$
  \item skew-$t$ marginal, $K=5$ knots, $T=-\infty$
  \item skew-$t$ marginal, $K=5$ knots, $T=q(0.80)$
\end{enumerate}
where $q(0.80)$ is the 80th sample quantile of the data.
The design matrix $\bX$ includes an the intercept with a prior of $\beta \sim \text{N}(0, 10)$.
The spatial covariance parameters have priors $\log(\nu) \sim \text{N}(-1.2, 1)$, $\gamma \sim \text{U}(0, 1)$, $\log(\rho) \sim \text{N}(-2, 1)$.
The skewness parameter has prior $\alpha \sim \text{N}(0, 2)$.
The residual variance terms have priors $\sigma^2_t(\bs) \sim \text{IG}(0.1, 0.1)$.

@@

\section{Data analysis}\label{s:analysis}
To illustrate this method, we consider the daily maximum 8-hour ozone measurements for July 2005 at 735 Air Quality System (AQS) monitoring sites in the eastern United States as the response.
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
Initially, we fit a linear regression assuming that the mean function can be expressed as
\begin{align}
  \mu_t(\bs) = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs). \label{eq:datamean}
\end{align}
The data from 5 July are shown in Figure \ref{fig:ozone} along with a Q-Q plot of the residuals compared to a skew-$t$ distribution with 8 d.f. and $\alpha = 1$.
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/ozone-5jul.pdf}
  \includegraphics[width=0.5\linewidth]{plots/qq-res.pdf}
  \caption{Ozone values on 5 July 2005 (left) Q-Q plot of the residuals (right)}
  \label{fig:ozone}
\end{figure}
\end{center}
The $\chi(h)$ and $\chi(t)$ plots indicate that the residuals still demonstrate spatial and temporal dependence (see Figure \ref{fig:chi-st})
\begin{center}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/chi-h-ozone.pdf}
  \includegraphics[width=0.5\linewidth]{plots/chi-t-ozone.pdf}
  \caption{$\chi(h)$ plot for the residuals. The vertical line indicates distance after which observations no longer demonstrate dependence (left) $\chi(t)$ plot for the residuals (right)}
  \label{fig:chi-st}
\end{figure}
\end{center}

\subsection{Model comparisons}
We fit the model using Gaussian and skew-$t$ marginal distributions, $K=1, 5, 10, 15$ partitions, with $Y(\bs)$ censored at $T = 0, 50, 75, 90$ ppb as described in Section \ref{s:data}.
We also include a max-stable analysis using the method by ????
All methods assume the mean function given in (\ref{eq:datamean}).
For each model, Brier scores and quantile scores were were averaged over all sites and days to obtain a single quantile score for each dataset.
At a particular threshold or quantile level, the model that fits the best is the one with the lowest score.
Looking at the heatplot in Figure \ref{fig:heatplot}, we can see that the methods using no thresholding tend to outperform those methods that do use thresholding.
\hl{Brian, I'm not sure that this is the best way to show this. Maybe a better heatplot would be quantile scores for $T=0$ for a few different quantiles. Then you can get a sense how the partitions impact the scores. The main problem is that $K=10, T=q(0.90)$ is much larger (5 - 7 times) than the rest of the quantile scores (across all values), so it really has an impact on the color scale}

\subsection{Results}\label{s:results}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{plots/heatplot.pdf}
  \includegraphics[width=0.5\linewidth]{plots/qscore-best.pdf}
  \caption{Heatplot of quantile scores for the 99th quantile (left) Quantile scores of Gaussian, skew-$t$ (K=10, T=q(0)), and Max-stable (right)}
  \label{fig:heatplot}
\end{figure}
